---
title: "Mountain_Climber_Death_Analysis"
author: "Shahan Ahmed"
date: "2024-02-11"
output: word_document
---
#Libraries:
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(viridis)
library(readr)
library(GGally)
library(survival)
```
#Set working directory
```{r}
setwd("S:/Portfolio Project")
```

#Importing dataset and cleaning
```{r}
Mountain_deaths <- read_csv("deaths_on_eight-thousanders.csv")

Mountain_deaths_clean <- na.omit(Mountain_deaths)

glimpse(Mountain_deaths_clean)
```


#Question:How has the number of climbing accidents changeed over the year? 

```{r}
accidents_by_year <- Mountain_deaths_clean %>%
  group_by(Date) %>%
  summarise(Count = n())



ggplot(accidents_by_year, aes(x = Date, y = Count)) +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "20 year", date_labels = "%Y") +  # Adjust intervals and labels
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Improve readability of x-axis labels
  labs(title = "Trend of Climbing Accidents Over the Years", x = "Year", y = "Number of Accidents")

```

#Question: What are the top five reasones for climbers death? 

```{r}
accidents_by_cause <- Mountain_deaths_clean %>%
  group_by(`Cause of death`) %>%
  summarise(Count = n())

#Filter for top 5 causes
top_accidents_by_cause <- accidents_by_cause %>%
  arrange(desc(Count)) %>%
  top_n(5, Count)

ggplot(top_accidents_by_cause, aes(x = reorder(`Cause of death`, Count), y = Count, fill = `Cause of death`)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        plot.margin = margin(10, 10, 10, 10)) +
  labs(title = "Top Five Climbing Accidents by Cause of Death",
       x = "Cause of Death",
       y = "Number of Accidents")

```


#Can we identify any correlations between climber nationality and the causes of death?


```{r}
Mountain_deaths_clean %>%
  group_by(Nationality) %>%
  summarise(Count = n()) %>%
  arrange(-Count) %>%
  head(n = 20)

top_nationalities_10 <- Mountain_deaths_clean %>%
  group_by(Nationality) %>%
  summarise(Count = n()) %>%
  arrange(-Count) %>%
  head(n = 10)

ggplot(top_nationalities_10, aes(x = reorder(Nationality, -Count), y = Count)) +
  geom_segment(aes(x = Nationality, xend = Nationality, y = 0, yend = Count), color = "grey") +
  geom_point(color = "steelblue", size = 5) +
  theme_minimal() +
  labs(title = "Top 10 Nationalities in Mountain Deaths",
       x = "Nationality", 
       y = "Count of Deaths") +
  coord_flip()
```

```{r}
# Create a contingency table and Chiq test

Death_by_Nationality <- table(Mountain_deaths_clean$Nationality, Mountain_deaths_clean$`Cause of death`)
Death_by_Nationality <- chisq.test(Death_by_Nationality)
print(Death_by_Nationality)
```
Implementing Monte Carlo Simulation:
This approach helps mitigate the issue with low expected counts in some cells of your contingency table and provides a more reliable p-value through simulation, especially suitable for large contingency tables with many degrees of freedom.

# Monte Carlo Simulation
```{r}
if(any(chi_res$expected < 5)) 
chi_res_mc <- chisq.test(Death_by_Nationality, simulate.p.value = TRUE, B = 2000)
print(chi_res_mc)
```
The results suggest there is a statistically significant association between climbers' nationalities and the causes of death recorded in the dataset. This means that the distribution of causes of death varies significantly across different nationalities.

```{r}

```


